---
title: "ddG and Viability Modeling"
author: "Yesol Sapozhnikov"
date: "2024-10-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(mrbglm)
library(ggpubr)
library(cowplot)
```

## Get data

```{r}
ddg_fold <- read.csv("../outputs/ddg_errors/phix_fold_err.csv")[,-1]
ddg_trimer <- read.csv("../outputs/ddg_errors/phix_trimer_err.csv")[,-1]
scores <- read_tsv("../outputs/enrich2_results/tsv/Gmut_growth_exp/main_identifiers_scores.tsv",
                   skip = 1)
names(scores)[1] <- "variant"
#Combine scores with ddg dataframes
ddg <- data.frame(variant = ddg_fold$mut,
                  ddg_fold = ddg_fold$total, ddg_trimer = ddg_trimer$total,
                  errbd_fold = ddg_fold$upr_bd, errbd_trimer = ddg_trimer$upr_bd,
                  errpt_fold = ddg_fold$err_pred, errpt_trimer = ddg_trimer$err_pred,
                  residue = ddg_fold$residue,
                  str = ddg_fold$str,
                  RSA = ddg_fold$RSA)
scores <- inner_join(scores, ddg) %>% arrange(variant)

# for counts data:
counts <- read.csv("../outputs/dada2result_yesol2.csv")
counts[counts$var == "wt",]$AAsub <- "_wt"
counts$AAsub <- gsub(",", "-", counts$AAsub) # change , with - for multiple AA subs for better readability
counts$sample <- str_replace(counts$sample, "lig.+", "lig") # combine all 3 reps of ligation mix since they're from same pool anyway.

# create missing counts category and simpler structure categories
inputs <- c("t0", 
            "t35_rep1", "t35_rep2", "t35_rep3",
            "t70_dil_day2_rep1", "t70_dil_day2_rep2", "t70_dil_day2_rep3")
vars_counts <- scores %>% dplyr::select(variant)

for (i in 1:length(inputs)){
  df <- counts %>% filter(sample == inputs[i]) %>% dplyr::select(AAsub, count)
  names(df) <- c("variant", inputs[i])
  vars_counts <- left_join(vars_counts, df)
}
vars_counts[is.na(vars_counts)] <- 0

# create category
for (i in 1:nrow(vars_counts)){
  t0_i <- vars_counts$t0[i]
  t35_i <- sum(vars_counts$t35_rep1[i], vars_counts$t35_rep2[i], vars_counts$t35_rep3[i])
  t70_i <- sum(vars_counts$t70_dil_day2_rep1[i], vars_counts$t70_dil_day2_rep2[i], vars_counts$t70_dil_day2_rep3[i])
  if (t0_i == 0 & t35_i == 0) vars_counts$missing[i] <-  "t0-t35"
  else if (t0_i == 0 & t70_i == 0) vars_counts$missing[i] <-  "t0-t70"
  else if (t0_i == 0) vars_counts$missing[i] <-  "t0"
  else if (t35_i == 0 & t70_i == 0) vars_counts$missing[i] <- "t35-t70"
  else if (t35_i == 0) vars_counts$missing[i] <- "t35"
  else if (t70_i == 0) vars_counts$missing[i] <- "t70"
  else vars_counts$missing[i] <- "none"
}
scores$missing <- as.factor(vars_counts$missing)

# remove some of the missing categories that don't appear to be contributive and 
# create a category of viable vs inviable

# all 688 observed singles
scores_viability_all <- scores %>% 
  mutate(viability = ifelse(missing == "t35-t70", 0, 1))
# filter out some missing categories 
scores_viability <- scores %>% filter(missing %in% c("none", "t0", "t35-t70", "t70")) %>%
  mutate(viability = ifelse(missing == "t35-t70", 0, 1))


########## Residue 119 ############
# Try removing all datapoints that are residue 119 based on what we know from SDM work.
# Total of 7 datapoints
scores_viability_all <- scores_viability_all %>% filter(residue != 119)
scores_viability <- scores_viability %>% filter(residue != 119)


######### Doubles ###########

scores_all <- read_tsv("../outputs/enrich2_results/tsv/Gmut_growth_exp/main_identifiers_scores.tsv",
                   skip = 1)
names(scores_all)[1] <- "variant"

scores_sing <- inner_join(scores_all, ddg) %>% arrange(variant)

# Create timepoint counts dataframe -- all scored variants
inputs <- c("t0", 
            "t35_rep1", "t35_rep2", "t35_rep3",
            "t70_dil_day2_rep1", "t70_dil_day2_rep2", "t70_dil_day2_rep3")
vars_counts2 <- scores_all %>% dplyr::select(variant)

for (i in 1:length(inputs)){
  df <- counts %>% filter(sample == inputs[i]) %>% dplyr::select(AAsub, count)
  names(df) <- c("variant", inputs[i])
  vars_counts2 <- left_join(vars_counts2, df)
}
vars_counts2[is.na(vars_counts2)] <- 0

# create category for missing timepoints
for (i in 1:nrow(vars_counts2)){
  t0_i <- vars_counts2$t0[i]
  t35_i <- sum(vars_counts2$t35_rep1[i], vars_counts2$t35_rep2[i], vars_counts2$t35_rep3[i])
  t70_i <- sum(vars_counts2$t70_dil_day2_rep1[i], vars_counts2$t70_dil_day2_rep2[i], vars_counts2$t70_dil_day2_rep3[i])
  if (t0_i == 0 & t35_i == 0) vars_counts2$missing[i] <-  "t0-t35"
  else if (t0_i == 0 & t70_i == 0) vars_counts2$missing[i] <-  "t0_t70"
  else if (t0_i == 0) vars_counts2$missing[i] <-  "t0"
  else if (t35_i == 0 & t70_i == 0) vars_counts2$missing[i] <- "t35-t70"
  else if (t35_i == 0) vars_counts2$missing[i] <- "t35"
  else if (t70_i == 0) vars_counts2$missing[i] <- "t70"
  else vars_counts2$missing[i] <- "none"
}
# create category for number of muts
vars_counts2 <- vars_counts2 %>%
  mutate(nmut = str_count(variant, "-") + 1,
         viability = ifelse(missing == "t35-t70", 0, 1))

# create a category of viable vs inviable
scores_viability <- left_join(scores_sing, vars_counts2[,c("variant", "missing")]) %>% 
  #filter(missing %in% c("none", "t0", "t35-t70", "t70")) %>%   ### omit this filtering for this analysis
  mutate(viability = ifelse(missing == "t35-t70", 0, 1))

# Create doubles dataframe containing time series counts, score, and constituent muts
doubles <- inner_join(scores_all, vars_counts2) %>% filter(nmut == 2) %>% select(-nmut)
# Add columns with each mut of the doubles
for (i in 1:nrow(doubles)){
  str <- doubles$variant[i]
  doubles$var1[i] <- str_split(str, "-")[[1]][1]
  doubles$var2[i] <- str_split(str, "-")[[1]][2]
}

counts <- 
  counts %>% filter(sample %in% inputs, nmut_tot <= 2) %>% 
    select(sample, AAsub, count, nmut_tot, nmut_offtarget)

vars_sing <- scores_sing$variant
doub <- data.frame(vars = doubles$variant,
                   mut1 = doubles$var1,
                   mut2 = doubles$var2,
                   mut1_present = rep(NA, nrow(doubles)),
                   mut2_present = rep(NA, nrow(doubles)),
                   both_present = rep(NA, nrow(doubles)))
for (i in 1:nrow(doub)){
  doub$mut1_present[i] <- doub$mut1[i] %in% vars_sing
  doub$mut2_present[i] <- doub$mut2[i] %in% vars_sing
  doub$both_present[i] <- doub$mut1_present[i] + doub$mut2_present[i]
}

doubles2 <- doubles %>% select(variant, score, viability:var2) %>%
  rename(mut1 = var1, mut2 = var2)
for (i in 1:nrow(doubles2)){
  mut <- doubles2$mut1[i] 
  if (str_sub(mut, nchar(mut), nchar(mut)) == "*"){
    doubles2$ddg_fold1[i] <- NA
    doubles2$ddg_trimer1[i] <- NA
  } else if (str_split(mut, "[0-9]+")[[1]][1] != str_split(mut, "[0-9]+")[[1]][2]){ # check for synonymous muts
    doubles2$ddg_fold1[i] <- ddg[ddg$variant == mut,]$ddg_fold
    doubles2$ddg_trimer1[i] <- ddg[ddg$variant == mut,]$ddg_trimer
  } else {
    doubles2$ddg_fold1[i] <- 0
    doubles2$ddg_trimer1[i] <- 0
  }
  if (doub$mut1_present[i] == TRUE){
    doubles2$score1[i] <- scores_viability[scores_viability$variant == mut,]$score
    doubles2$viab1[i] <- scores_viability[scores_viability$variant == mut,]$viability
    doubles2$as_sing1[i] <- TRUE
  } else {
    doubles2$score1[i] <- NA
    doubles2$viab1[i] <- NA
    doubles2$as_sing1[i] <- FALSE
  }
  mut <- doubles2$mut2[i] 
  if (str_sub(mut, nchar(mut), nchar(mut)) == "*"){
    doubles2$ddg_fold2[i] <- NA
    doubles2$ddg_trimer2[i] <- NA
  } else if (str_split(mut, "[0-9]+")[[1]][1] != str_split(mut, "[0-9]+")[[1]][2]){
    doubles2$ddg_fold2[i] <- ddg[ddg$variant == mut,]$ddg_fold
    doubles2$ddg_trimer2[i] <- ddg[ddg$variant == mut,]$ddg_trimer
  } else {
    doubles2$ddg_fold2[i] <- 0
    doubles2$ddg_trimer2[i] <- 0
  }
  if (doub$mut2_present[i] == TRUE){
    doubles2$score2[i] <- scores_viability[scores_viability$variant == mut,]$score
    doubles2$viab2[i] <- scores_viability[scores_viability$variant == mut,]$viability
    doubles2$as_sing2[i] <- TRUE
  } else {
    doubles2$score2[i] <- NA
    doubles2$viab2[i] <- NA
    doubles2$as_sing2[i] <- FALSE
  }
}

doubles2 <- doubles2 %>%
  mutate(ddg_fold_sum = ddg_fold1 + ddg_fold2,
         ddg_trimer_sum = ddg_trimer1 + ddg_trimer2)

doubles2 <- doubles2 %>% filter(!is.na(ddg_fold_sum), !is.na(ddg_trimer_sum))


viab_double <- doubles2 %>% 
  select(variant, viability) %>%
  mutate(type = "double")
viab_single <- scores_viability_all %>%
  select(variant, viability) %>%
  mutate(type = "single")
viab_comb <- bind_rows(viab_single, viab_double)
viab_df <- viab_comb %>% group_by(type, viability) %>%
  summarise(unique_vars = n()) %>%
  mutate(viability = factor(viability),
         type = factor(type, levels = c("single", "double")))

```


Plot # viable and inviable as stacked bar chart, histograms with ddG.
```{r}
bar_viab_variants <-
ggplot(viab_df) +
  geom_bar(aes(x = type, y = unique_vars, fill = viability),
           position = "stack", stat = "identity", width = .3) +
  scale_fill_manual(values = c("darkorange2", "royalblue4"),
                    name = "Viability",
                    labels = c("inviable", "viable")) +
  labs(y = "Number of unique variants") +
  scale_x_discrete(labels = c("single" = "Singles", "double" = "Doubles")) +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        #legend.position = c(0.2, 0.85)
  )
hist_vars_along_res <-
scores_viability_all %>%
  ggplot(aes(x = residue)) +
  geom_histogram(aes(fill = as.factor(viability)), binwidth = 1) +
  scale_fill_manual(values = c("darkorange2", "royalblue4"),
                    name = "Viability",
                    labels = c("inviable", "viable")) +
  labs(x = "Mutated residue", y = "Unique variants") +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_text(size = 10))
ddg_along_res <- 
scores_viability_all %>%
  ggplot(aes(x = residue, y = ddg_fold)) +
  geom_point(aes(color = as.factor(viability))) +
  scale_color_manual(values = c("darkorange2", "royalblue4"), name = "Viability",
                    labels = c("inviable", "viable")) +
  labs(x = "Mutated residue", y = expression(paste(Delta, Delta, "G")[fold]~(kcal/mol))) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_text(size = 10))
```

```{r}
ggarrange(bar_viab_variants,
          ggarrange(hist_vars_along_res, ddg_along_res,
                    ncol = 1, labels = c("B", "C")),
          ncol = 2,
          labels = "A")

# ggsave("../figures/overview_plots.pdf",
#       width = 200, height = 100, units = "mm", dpi = 300)
```


## Viability modeling - Singles

MBM fitting

Asymptote estimate
```{r}
## quantile restrictions on the regression intercepts (an intercept is 
#  restricted to the interval [lowxq, uppxq] of the corresponding
#  predictor); defaults to [-Inf, Inf].
lowxq <- 'xmin'
uppxq <- 'xq95' # we would not want an intercept to be in the right tail of predictor distribution

#Simple r-squared values between y and each predictor x
cor(cbind(scores_viability_all[, c('viability', 'ddg_fold', 'ddg_trimer')]))[1, 2:3]^2

# Because we have more than one predictor, we estimate the asymptote using the empirical method (as opposed to maximum likelihood)

# Crude estimate
Linf_single <- infer.lambda(y = scores_viability_all$viability, x = scores_viability_all[,c('ddg_fold', 'ddg_trimer')])

# The crude asymptote estimate, and its 95% confidence interval 
Linf_single$lambda
Linf_single$conf.int

# Correct the empirical asymptote estimator (using a naive ML estimate of the regression coefficients)
mu_ab <- Linf_single$lambda
pick <- Linf_single$which
npick <- length(pick)

single_no_err0 <- glm.mrb(viability ~ ddg_fold + ddg_trimer, data = scores_viability_all,
                         control = list(fixL = FALSE,
                                        lower.intercepts = lowxq,
                                        upper.intercepts = uppxq,
                                        penalize.intercepts = FALSE),
                         Lstart = Linf_single$lambda,
                         x = TRUE, y = TRUE)

hvalues <- single_no_err0$mu[pick] / single_no_err0$L.values
omega_ab <- mean(hvalues)
omega2_ab <- var(hvalues) / npick
Lhat <- max(min(mu_ab / (omega_ab - omega2_ab), 1-single_no_err0$control$epsilon), single_no_err0$control$epsilon)
Lconf.int <- pmax(pmin(Linf_single$conf.int / (omega_ab - omega2_ab), 1-single_no_err0$control$epsilon), single_no_err0$control$epsilon) # Assuming omega_ab is a constant

# Final estimates and 95% confidence interval 
Lhat
Lconf.int
```

# Fit without measurement error
```{r }
# Fit and summarize the model
single_no_err <- glm.mrb(viability ~ ddg_fold + ddg_trimer, data = scores_viability_all,
                         control = list(fixL = TRUE,
                                        lower.intercepts = lowxq,
                                        upper.intercepts = uppxq,
                                        penalize.intercepts = FALSE,
                                        nobs.power = 0.5),
                         Lstart = Lhat,
                         x = TRUE, y = TRUE)
summary(single_no_err)

```



# Fit with measurement error

```{r}
# Correct the empirical asymptote estimate (using a naive ML estimate of the regression coefficients)
mu_ab <- Linf_single$lambda
pick <- Linf_single$which
npick <- length(pick)

single_err0 <- glm.mrb(viability~ddg_fold+ddg_trimer, data = scores_viability_all,
                      x.with.me.offsets = c(TRUE, TRUE),
                      me.offsets = ~(abs(errpt_fold))+(abs(errpt_trimer)),
                      minp.formula = ~-1,
                      maxp.formula = ~1,
                      control = list(fixL = FALSE,
                                        lower.intercepts = lowxq,
                                        upper.intercepts = uppxq,
                                        penalize.intercepts = FALSE),
                         Lstart = Linf_single$lambda,
                      link = 'logit')
summary(single_err0)

hvalues <- single_err0$mu[pick] / single_err0$L.values
omega_ab <- mean(hvalues)
omega2_ab <- var(hvalues) / npick
Lhat <- max(min(mu_ab / (omega_ab - omega2_ab), 1-single_err0$control$epsilon), single_err0$control$epsilon)
Lconf.int <- pmax(pmin(Linf_single$conf.int / (omega_ab - omega2_ab), 1-single_err0$control$epsilon), single_err0$control$epsilon) # Assuming omega_ab is a constant

# Final asymptote estimates and 95% confidence interval 
Lhat
Lconf.int
```

```{r }
single_err <- glm.mrb(viability~ddg_fold+ddg_trimer, data = scores_viability_all,
                      x.with.me.offsets = c(TRUE, TRUE),
                      me.offsets = ~(abs(errpt_fold))+(abs(errpt_trimer)),
                      minp.formula = ~-1,
                      maxp.formula = ~1,
                      control = list(fixL = TRUE,
                                     lower.intercepts = lowxq,
                                     upper.intercepts = uppxq,
                                     nobs.power = 0.5,
                                     slope.signs = "negative"),
                         Lstart = Lhat,
                      link = 'logit')
summary(single_err)
```

Contour plot and individual probability curves

```{r}
# Create a new df to store predicted values
viab_pred <- scores_viability_all %>%
  select(variant, ddg_fold, ddg_trimer, errpt_fold, errpt_trimer, viability)
x <- seq(-3, max(scores_viability_all$ddg_fold), length.out = 100)
y <- seq(-3, 20, length.out = 100)
grid <- expand.grid(x, y)
viab_pred <- data.frame(ddg_fold = grid$Var1,
                        ddg_trimer = grid$Var2)
viab_pred$bi_fit <- predict(single_err, newdata = viab_pred)
viab_pred$fold_fit <- predict(single_err,
                              newdata = data.frame(ddg_fold = viab_pred$ddg_fold,
                                                   ddg_trimer = rep(0, nrow(viab_pred))))
viab_pred$bind_fit <- predict(single_err,
                              newdata = data.frame(ddg_fold = rep(0, nrow(viab_pred)),
                                                   ddg_trimer = viab_pred$ddg_trimer))

pal <- colorRampPalette(c("#e9ede1", "#2dada3", "#20488a"))
colors10 <- pal(10)

# Contour plot
p1 <- 
ggplot(viab_pred, aes(ddg_fold, ddg_trimer)) +
  # contour plot (legend removed because discrete legend doesn't look good)
  geom_contour_filled(aes(z = bi_fit)) +
  scale_fill_manual(values = colors10, 
                    guide = "none") +
  coord_cartesian(xlim = c(-3, 20), ylim = c(-3, 20)) +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +
  # create points (that will be covered up by the second geom_point) for the binned color legend
  geom_point(data = scores_viability_all, aes(ddg_fold, ddg_trimer, 
                                          color = single_err$fitted.values, shape = factor(viability))) +
  scale_color_stepsn(n.breaks = 10, colors = colors10) +
  # create points to show viable/inviable datapoints
  geom_point(data = scores_viability_all, aes(ddg_fold, ddg_trimer, shape = factor(viability)),
             size = 2) +
  scale_shape_manual(values = c(4, 20)) +
  
  labs(color = "Probability", shape = "Viability",
       x = expression(paste(Delta, Delta, "G")[fold]~(kcal/mol)),
       y = expression(paste(Delta, Delta, "G")[bind]~(kcal/mol))) +
  theme(legend.direction = "vertical",
        legend.box = "horizontal",
        legend.position = c(.7, .7),
        axis.title = element_text(size = 12), axis.text = element_text(size = 10),
        legend.spacing.y = unit(0.1, "cm"),
        legend.text = element_text(size = 8),
        plot.margin = margin(t = 10, r = 20, b = 10, l = 10))

# probability curve while ddg_bind is held at zero
start <- min(scores_viability_all$ddg_fold)
end <- max(scores_viability_all$ddg_fold)
x <- seq(start, end, by = 0.05)

prob <- predict(single_err,
             newdata = data.frame(
               ddg_fold = x,
               ddg_trimer = rep(0, length(x))
             ))
p2 <- 
ggplot(scores_viability_all, aes(x = ddg_fold, y = viability)) +
  geom_jitter(height = 0.15, color = "darkgrey", size = 0.5) +
  theme_classic() +
  geom_line(data = data.frame(x = x, y = prob),
             aes(x = x, y = y), linewidth = 1.5) +
  labs(x = expression(paste(Delta, Delta, "G")[fold]~(kcal/mol)),
       y = "Viability") +
  theme(axis.title = element_text(size = 12), axis.text = element_text(size = 10)) +
  scale_x_continuous(limits = c(-4, 33)) +
  scale_y_continuous(breaks = c(0, 1), limits = c(-.2, 1.2))

# probability curve while ddg_fold is held at zero
start <- min(scores_viability_all$ddg_trimer)
end <- max(scores_viability_all$ddg_fold)
x <- seq(start, end, by = 0.05)

prob <- predict(single_err,
             newdata = data.frame(
               ddg_fold = rep(0, length(x)),
               ddg_trimer = x
             ))
p3 <- 
ggplot(scores_viability_all, aes(x = ddg_trimer, y = viability)) +
  geom_jitter(height = 0.15, color = "darkgrey", size = 0.5) +
  theme_classic() +
  geom_line(data = data.frame(x = x, y = prob),
             aes(x = x, y = y), linewidth = 1.5) +
  labs(x = expression(paste(Delta, Delta, "G")[bind]~(kcal/mol)),
       y = "Viability") +
  theme(axis.title = element_text(size = 12), axis.text = element_text(size = 10)) +
  scale_x_continuous(limits = c(-4, 33)) +
  scale_y_continuous(breaks = c(0, 1), limits = c(-.2, 1.2))

# Put all 3 plots together
right_col <- plot_grid(p2, p3, nrow = 2)
plot_grid(p1, right_col, ncol = 2, rel_widths = c(3, 2), labels = c("A", "B"))
#ggsave("../figures/viab_models.pdf", width = 170, height = 90, units = "mm", dpi = 300)
```



## Linear regression modeling

```{r}
# Take viable variants only, also add max ddG.
scores2 <- scores_viability_all %>% filter(viability == 1) %>%
  mutate(max_ddg = pmax(ddg_fold, ddg_trimer))
lm_modw <- lm(score~ddg_fold+ddg_trimer, data = scores2, weights = 1/(errbd_fold+errbd_trimer))
summary(lm_modw)
```
Plot the effect of each ddg separately
```{r}
# Fit 
lm_f <- lm(score~ddg_fold, data = scores2, weights = 1/errbd_fold)
lm_b <- lm(score~ddg_trimer, data = scores2, weight = 1/errbd_trimer)

# get lm_f coef and plot
eq <- paste0("y = ", round(lm_f$coefficients[2], 2), "x + ", round(lm_f$coefficients[1], 2))
score_fold <- 
ggplot(scores2, aes(ddg_fold, score)) +
  geom_point(size = .8) + #geom_smooth(method = "lm", se = F) +
  geom_abline(intercept = lm_f$coefficients[1], slope = lm_f$coefficients[2], color = "red") +
  labs(
    x = expression(paste(Delta, Delta, "G")[fold]~(kcal/mol)), y = "Score"
    ) +
  theme_bw() + theme(axis.text = element_text(size = 12),
                     axis.title = element_text(size = 14)) +
  scale_x_continuous(limits = c(-3, 10)) +
  annotate("text", label = eq, x = 7, y = -5, size = 4) +
  annotate("text", x = 7, y = -6, size = 4, 
           label = "paste(italic(R) ^ 2, \" = 0.01\")", parse = TRUE)

# get lm_b coef and plot
eq <- paste0("y = ", round(lm_b$coefficients[2], 2), "x + ", round(lm_b$coefficients[1], 2))
score_bind <- 
ggplot(scores2, aes(ddg_trimer, score)) +
  geom_point(size = .8) + 
  geom_abline(intercept = lm_b$coefficients[1], slope = lm_b$coefficients[2], color = "red") +
  labs(
    x = expression(paste(Delta, Delta, "G")[bind]~(kcal/mol)), y = NULL
    ) +
  theme_bw() + theme(axis.text = element_text(size = 12),
                     axis.title = element_text(size = 14)) +
  scale_x_continuous(limits = c(-3, 10)) +
  annotate("text", label = eq, x = 7, y = -5, size = 4) +
  annotate("text", x = 7, y = -6, size = 4, 
           label = "paste(italic(R) ^ 2, \" = 0.003\")", parse = TRUE)

# combine both
score_both <- 
plot_grid(score_fold, score_bind, labels = c("A", "B"))
# ggsave("../figures/scores.pdf",
#        width = 170, height = 90, units = "mm", dpi = 300)
```

## Viability modeling - Doubles

```{r}
doubles2_err <- doubles2 %>% select(variant, viability, mut1, mut2, ddg_fold_sum, ddg_trimer_sum)
for (i in 1:nrow(doubles2_err)){
  err_fold1 <- abs(ddg[ddg$variant == doubles2_err$mut1[i], "errpt_fold"])
  err_fold2 <- abs(ddg[ddg$variant == doubles2_err$mut2[i], "errpt_fold"])
  err_trimer1 <- abs(ddg[ddg$variant == doubles2_err$mut1[i], "errpt_trimer"])
  err_trimer2 <- abs(ddg[ddg$variant == doubles2_err$mut2[i], "errpt_trimer"])
  doubles2_err$errpt_fold_sum[i] <- sum(err_fold1, err_fold2)
  doubles2_err$errpt_trimer_sum[i] <- sum(err_trimer1, err_trimer2)
}


# Because we have more than one predictor, we estimate the asymptote using the empirical method (as opposed to maximum likelihood)

# Crude estimate
Linf_double <- infer.lambda(y = doubles2_err$viability, x = doubles2_err[,c('ddg_fold_sum', 'ddg_trimer_sum')])

# The crude asymptote estimate, and its 95% confidence interval 
Linf_double$lambda
Linf_double$conf.int

# Correct the empirical asymptote estimator (using a naive ML estimate of the regression coefficients)
mu_ab <- Linf_double$lambda
pick <- Linf_double$which
npick <- length(pick)

double_no_err0 <- glm.mrb(viability ~ ddg_fold_sum + ddg_trimer_sum, data = doubles2_err,
                         control = list(fixL = FALSE,
                                        lower.intercepts = lowxq,
                                        upper.intercepts = uppxq,
                                        penalize.intercepts = FALSE),
                         Lstart = Linf_double$lambda,
                         x = TRUE, y = TRUE)

hvalues <- double_no_err0$mu[pick] / double_no_err0$L.values
omega_ab <- mean(hvalues)
omega2_ab <- var(hvalues) / npick
Lhat <- max(min(mu_ab / (omega_ab - omega2_ab), 1-double_no_err0$control$epsilon), double_no_err0$control$epsilon)
Lconf.int <- pmax(pmin(Linf_double$conf.int / (omega_ab - omega2_ab), 1-double_no_err0$control$epsilon), double_no_err0$control$epsilon) # Assuming omega_ab is a constant

# Final estimates and 95% confidence interval 
Lhat
Lconf.int

# Fit and summarize the model
double_no_err <- glm.mrb(viability~ddg_fold_sum+ddg_trimer_sum, data = doubles2_err,
                         control = list(fixL = TRUE,
                                        lower.intercepts = lowxq,
                                        upper.intercepts = uppxq,
                                        penalize.intercepts = FALSE,
                                        nobs.power = 0.5),
                         Lstart = Lhat,
                         x = TRUE, y = TRUE)
summary(double_no_err)
```

With measurement error
```{r}
# Correct the empirical asymptote estimator (using a naive ML estimate of the regression coefficients)
mu_ab <- Linf_double$lambda
pick <- Linf_double$which
npick <- length(pick)

double_err0 <- glm.mrb(viability ~ ddg_fold_sum + ddg_trimer_sum, data = doubles2_err,
                       x.with.me.offsets = c(TRUE, TRUE),
                      me.offsets = ~(abs(errpt_fold_sum))+(abs(errpt_trimer_sum)),
                      minp.formula = ~-1,
                      maxp.formula = ~1,
                         control = list(fixL = FALSE,
                                        lower.intercepts = lowxq,
                                        upper.intercepts = uppxq,
                                        penalize.intercepts = FALSE),
                         Lstart = Linf_double$lambda,
                         x = TRUE, y = TRUE)

hvalues <- double_err0$mu[pick] / double_err0$L.values
omega_ab <- mean(hvalues)
omega2_ab <- var(hvalues) / npick
Lhat <- max(min(mu_ab / (omega_ab - omega2_ab), 1-double_err0$control$epsilon), double_err0$control$epsilon)
Lconf.int <- pmax(pmin(Linf_double$conf.int / (omega_ab - omega2_ab), 1-double_err0$control$epsilon), double_err0$control$epsilon) # Assuming omega_ab is a constant

# Final estimates and 95% confidence interval 
Lhat
Lconf.int

# Fit
double_err <- glm.mrb(viability~ddg_fold_sum+ddg_trimer_sum, data = doubles2_err,
                     x.with.me.offsets = c(TRUE, TRUE),
                      me.offsets = ~(abs(errpt_fold_sum))+(abs(errpt_trimer_sum)),
                      minp.formula = ~-1,
                      maxp.formula = ~1,
                      control = list(fixL = TRUE,
                                     lower.intercepts = lowxq,
                                     upper.intercepts = uppxq,
                                     nobs.power = 0.5),
                      Lstart = Lhat,
                      link = 'logit')
summary(double_err)
```

Figures
```{r}
# bivariate contour plot with scatter plot overlay
x <- seq(-3, max(doubles2_err$ddg_fold_sum), length.out = 100)
y <- seq(-3, 20, length.out = 100)
grid <- expand.grid(x, y)
doubles2_err_pred <- data.frame(ddg_fold_sum = grid$Var1,
                                ddg_trimer_sum = grid$Var2)
doubles2_err_pred$bi_fit <- predict(double_err,
                                    newdata = doubles2_err_pred)
doubles2_err_pred$fold_fit <- 
  predict(double_err, 
          newdata = data.frame(ddg_fold_sum = doubles2_err_pred$ddg_fold_sum, 
                               ddg_trimer_sum = rep(0, nrow(doubles2_err_pred))))
doubles2_err_pred$bind_fit <- 
  predict(double_err,
          newdata = data.frame(ddg_fold_sum = rep(0, nrow(doubles2_err_pred)),
                               ddg_trimer_sum = doubles2_err_pred$ddg_trimer_sum))
pal <- colorRampPalette(c("#e9ede1", "#2dada3", "#20488a"))
colors10 <- pal(10)
p1 <-
ggplot(doubles2_err_pred, aes(ddg_fold_sum, ddg_trimer_sum)) +
  # contour plot (legend removed because discrete legend doesn't look good)
  geom_contour_filled(aes(z = bi_fit)) +
  scale_fill_manual(values = colors10, 
                    guide = "none") +
  coord_cartesian(xlim = c(-3, 20), ylim = c(-3, 20)) +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +
  # create points (that will be covered up by the second geom_point) for the binned color legend
  geom_point(data = doubles2_err, aes(ddg_fold_sum, ddg_trimer_sum, 
                                          color = double_err$fitted.values, shape = factor(viability))) +
  scale_color_stepsn(n.breaks = 10, colors = colors10) +
  # create points to show viable/inviable datapoints
  geom_point(data = doubles2_err, aes(ddg_fold_sum, ddg_trimer_sum, shape = factor(viability)),
             size = 2) +
  scale_shape_manual(values = c(4, 20)) +
  
  labs(color = "Probability", shape = "Viability",
       x = expression(paste("Sum of ", Delta, Delta, "G")[fold]~(kcal/mol)),
       y = expression(paste("Sum of ", Delta, Delta, "G")[bind]~(kcal/mol))) +
  theme(legend.direction = "vertical",
        legend.box = "horizontal",
        legend.position = c(.7, .7),
        axis.title = element_text(size = 12), axis.text = element_text(size = 10),
        legend.spacing.y = unit(0.1, "cm"),
        legend.text = element_text(size = 8),
        plot.margin = margin(t = 10, r = 20, b = 10, l = 10))

# visualize ddg_fold_sum effect on probability while ddg_bind held at zero
start <- min(doubles2_err$ddg_fold_sum)
end <- max(doubles2_err$ddg_fold_sum)
x <- seq(start, end, by = 0.05)
prob <- predict(double_err, newdata = data.frame(ddg_fold_sum = x,
                                         ddg_trimer_sum = rep(0, length(x))))
p2 <- 
ggplot(doubles2_err, aes(ddg_fold_sum, y = viability)) +
  geom_jitter(height = 0.15, color = "darkgrey", size = 0.5) +
  theme_classic() +
  geom_line(data = data.frame(x = x, y = prob),
            aes(x = x, y = y), linewidth = 1.5) +
  labs(x = expression(paste("Sum of ", Delta, Delta, "G")[fold]~(kcal/mol)),
       y = "Viability") +
  theme(axis.title = element_text(size = 12), axis.text = element_text(size = 10)) +
  scale_x_continuous(limits = c(-4, 35)) +
  scale_y_continuous(breaks = c(0, 1), limits = c(-.2, 1.2))

# visualize ddg_trimer_sum effect while ddg_fold held at zero
start <- min(doubles2_err$ddg_trimer_sum)
end <- max(doubles2_err$ddg_fold_sum)
x <- seq(start, end, by = 0.05)
prob <- predict(double_err, newdata = data.frame(ddg_fold_sum = rep(0, length(x)),
                                         ddg_trimer_sum = x))
p3 <- 
ggplot(doubles2_err, aes(x = ddg_trimer_sum, y = viability)) +
  geom_jitter(height = 0.15, color = "darkgrey", size = 0.5) +
  theme_classic() +
  geom_line(data = data.frame(x = x, y = prob),
             aes(x = x, y = y), linewidth = 1.5) +
  labs(x = expression(paste("Sum of ", Delta, Delta, "G")[bind]~(kcal/mol)),
       y = "Viability") +
  theme(axis.title = element_text(size = 12), axis.text = element_text(size = 10)) +
  scale_x_continuous(limits = c(-4, 33)) +
  scale_y_continuous(breaks = c(0, 1), limits = c(-.2, 1.2))

# Put 3 plots together
right_col <- plot_grid(p2, p3, nrow = 2)
plot_grid(p1, right_col, ncol = 2, rel_widths = c(3,2), labels = c("A", "B"))

#ggsave("../figures/doub_viab_models.pdf", width = 170, height = 90, units = "mm", dpi = 300)
```



## Doubles viability relationship with singles viability

```{r}
# Create a column indicating how many of the constituent muts were observed in singles data
for (i in 1:nrow(doubles2)){
  if (doubles2$as_sing1[i] == TRUE & doubles2$as_sing2[i] == TRUE){
    doubles2$n_singles_data[i] <- 2
  } else if (doubles2$as_sing1[i] == TRUE | doubles2$as_sing2[i] == TRUE){
    doubles2$n_singles_data[i] <- 1
  } else {
    doubles2$n_singles_data[i] <- 0
  }
}
table(doubles2$n_singles_data)
```

```{r}
# dataframe for doubles with both known singles
doubles2_two <- doubles2 %>% filter(n_singles_data == 2) %>%
  select(variant, viability:ddg_trimer1, viab1, 
         ddg_fold2:ddg_trimer2, viab2, ddg_fold_sum, ddg_trimer_sum) %>%
  rename(mut1_viab = viab1, mut2_viab = viab2, double_viab = viability)

for (i in 1:nrow(doubles2_two)){
  if (doubles2_two$mut1_viab[i] == TRUE && doubles2_two$mut2_viab[i] == TRUE){
    doubles2_two$singles_info[i] <- "both viable"
  }
  else if (doubles2_two$mut1_viab[i] == FALSE && doubles2_two$mut2_viab[i] == FALSE){
    doubles2_two$singles_info[i] <- "both inviable"
  } else {
    doubles2_two$singles_info[i] <- "mixed"
  }
}

# dataframe for doubles with only one of the muts known as a single
doubles2_one <- doubles2 %>% filter(n_singles_data == 1)
doubles2_one2 <- tibble(variant = doubles2_one$variant,
                        ddg_fold_sum = doubles2_one$ddg_fold_sum,
                        ddg_trimer_sum = doubles2_one$ddg_trimer_sum)
for (i in 1:nrow(doubles2_one)){
  if (doubles2_one$as_sing1[i] == TRUE){
    doubles2_one2$mut[i] <- doubles2_one$mut1[i]
    doubles2_one2$double_viab[i] <- doubles2_one$viability[i]
    doubles2_one2$mut_viab[i] <- doubles2_one$viab1[i]
    doubles2_one2$ddg_fold[i] <- doubles2_one$ddg_fold1[i]
    doubles2_one2$ddg_trimer[i] <- doubles2_one$ddg_trimer1[i]
    doubles2_one2$ddg_fold_other[i] <- doubles2_one$ddg_fold2[i]
    doubles2_one2$ddg_trimer_other[i] <- doubles2_one$ddg_trimer2[i]
  } else{
    doubles2_one2$mut[i] <- doubles2_one$mut2[i]
    doubles2_one2$double_viab[i] <- doubles2_one$viability[i]
    doubles2_one2$mut_viab[i] <- doubles2_one$viab2[i]
    doubles2_one2$ddg_fold[i] <- doubles2_one$ddg_fold2[i]
    doubles2_one2$ddg_trimer[i] <- doubles2_one$ddg_trimer2[i]
    doubles2_one2$ddg_fold_other[i] <- doubles2_one$ddg_fold1[i]
    doubles2_one2$ddg_trimer_other[i] <- doubles2_one$ddg_trimer1[i]
  }
} 

doubles2_one2$mut_viab_label <- 
  ifelse(doubles2_one2$mut_viab == 1, "one mut viable", "one mut inviable")

doubles2_one2 %>% group_by(mut_viab_label, double_viab) %>% count()

```
Statistical tests
```{r}
doubles2_two$total_sing_viab <- doubles2_two$mut1_viab + doubles2_two$mut2_viab
table(doubles2_two$double_viab, doubles2_two$total_sing_viab)
chisq.test(table(doubles2_two$double_viab, doubles2_two$total_sing_viab))

table(doubles2_one2$double_viab, doubles2_one2$mut_viab_label)
chisq.test(table(doubles2_one2$double_viab, doubles2_one2$mut_viab_label))
```
Plots
```{r}
df1 <- doubles2_two %>% filter(total_sing_viab == 2) %>% select(variant, double_viab) %>% 
  mutate(inviab_mut = "none", inviab_mut2 = "Both viable")
df2 <- doubles2_two %>% filter(total_sing_viab == 1) %>% select(variant, double_viab) %>% 
  mutate(inviab_mut = "one", inviab_mut2 = "At least one inviable")
df3 <- doubles2_two %>% filter(total_sing_viab == 0) %>% select(variant, double_viab) %>% 
  mutate(inviab_mut = "both", inviab_mut2 = "At least one inviable")
df4 <- doubles2_one2 %>% filter(mut_viab_label == "one mut inviable") %>% select(variant, double_viab) %>%
  mutate(inviab_mut = "info_missing", inviab_mut2 = "At least one inviable")
df <- bind_rows(df1, df2, df3, df4)
df$double_viab <- factor(df$double_viab)
df$inviab_mut <- factor(df$inviab_mut, levels = c("none", "one", "both", "info_missing"))
df$inviab_mut2 <- factor(df$inviab_mut2, levels = c("Both viable", "At least one inviable"))
table(df$double_viab, df$inviab_mut)
table(df$double_viab, df$inviab_mut2)

label_inviab <- data.frame(label = c("inviable = 2", "inviable = 22", "inviable = 21"),
                         singles_info = c("both inviable", "both viable", "mixed"),
                         x = c(25, 25, 25), y = c(6, 6, 6)) 
label_viab <- data.frame(label = c("viable = 0", "viable = 214", "viable = 7"),
                         singles_info = c("both inviable", "both viable", "mixed"),
                         x = c(25, 25, 25), y = c(4.5, 4.5, 4.5)) 
p_doub_bar <-
df %>%
  ggplot(aes(x = inviab_mut2, fill = double_viab)) +
    geom_bar(stat = "count", width = 0.4) +
    scale_fill_manual(values = c("darkorange2", "royalblue4"),
                      name = "Double's viability",
                      labels = c("inviable", "viable")) +
    labs(y = "Number of unique variants") +
    theme_bw() +
    theme(axis.title.x = element_blank(),
         legend.position = c(0.7, 0.85))
# repeat of previous plot but change color and remove legend
p_doub1_2 <- 
doubles2_two %>% ggplot(aes(ddg_fold_sum, ddg_trimer_sum)) +
  geom_point(aes(color = as.factor(double_viab)), size = 0.7) +
  scale_color_manual(values = c("darkorange2", "royalblue4"),
                     name = "Double's viability",
                     labels = c("inviable", "viable")) +
  facet_grid(rows = vars(singles_info)) +
  theme_bw() +
  geom_text(data = label_inviab,
            aes(x = x, y = y, label = label), size = 3) +
  geom_text(data = label_viab,
            aes(x = x, y = y, label = label), size = 3) +
  labs(x = expression(paste("Sum of ", Delta, Delta, "G")[fold]~(kcal/mol)),
       y = expression(paste("Sum of ", Delta, Delta, "G")[bind]~(kcal/mol))) +
  theme(legend.position = "null",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 8))

plot_grid(p_doub_bar, p_doub1_2, labels = c("A", "B"),
          rel_widths = c(.8, 1))
ggsave("../figures/doub_viability2.pdf", width = 200, height = 90, units = "mm", dpi = 300)
```



